---
title: "Analyzing Data from Within-Subjects Designs: Multivariate Approach vs. Linear Mixed Models Approach"
output: 
  html_document: 
    theme: readable
---

# packages

```{r, warning = FALSE, message = FALSE}

# install.packages("tidyverse")
# install.packages("knitr")
# install.packages("lattice")
# install.packages("AMCP")
# install.packages("lme4")
# install.packages("lmerTest")

library(tidyverse)
library(knitr)
library(lattice)
library(AMCP)
library(lme4)
library(lmerTest)

```

# custom function

```{r}

oneway <- function(dv, group, contrast, alpha = .05) {
  # -- arguments --
  # dv: vector of measurements (i.e., dependent variable)
  # group: vector that identifies which group the dv measurement came from
  # contrast: list of named contrasts
  # alpha: alpha level for 1 - alpha confidence level
  # -- output --
  # computes confidence interval and test statistic for a linear contrast of population means in a between-subjects design
  # returns a data.frame object
  # estimate (est), standard error (se), t-statistic (z), degrees of freedom (df), two-tailed p-value (p), and lower (lwr) and upper (upr) confidence limits at requested 1 - alpha confidence level
  # first line reports test statistics that assume variances are equal
  # second line reports test statistics that do not assume variances are equal
  
  # means, standard deviations, and sample sizes
  ms <- by(dv, group, mean, na.rm = TRUE)
  vars <- by(dv, group, var, na.rm = TRUE)
  ns <- by(dv, group, function(x) sum(!is.na(x)))
  
  # convert list of contrasts to a matrix of named contrasts by row
  contrast <- matrix(unlist(contrast), nrow = length(contrast), byrow = TRUE, dimnames = list(names(contrast), NULL))
  
  # contrast estimate
  est <- contrast %*% ms
  
  # welch test statistic
  se_welch <- sqrt(contrast^2 %*% (vars / ns))
  t_welch <- est / se_welch
  
  # classic test statistic
  mse <- anova(lm(dv ~ factor(group)))$"Mean Sq"[2]
  se_classic <- sqrt(mse * (contrast^2 %*% (1 / ns)))
  t_classic <- est / se_classic
  
  # if dimensions of contrast are NULL, nummer of contrasts = 1, if not, nummer of contrasts = dimensions of contrast
  num_contrast <- ifelse(is.null(dim(contrast)), 1, dim(contrast)[1])
  df_welch <- rep(0, num_contrast)
  df_classic <- rep(0, num_contrast)
  
  # makes rows of contrasts if contrast dimensions aren't NULL
  if(is.null(dim(contrast))) contrast <- t(as.matrix(contrast))
  
  # calculating degrees of freedom for welch and classic
  for(i in 1:num_contrast) {
    df_classic[i] <- sum(ns) - length(ns)
    df_welch[i] <- sum(contrast[i, ]^2 * vars / ns)^2 / sum((contrast[i, ]^2 * vars / ns)^2 / (ns - 1))
  }
  
  # p-values
  p_welch <- 2 * (1 - pt(abs(t_welch), df_welch))
  p_classic <- 2 * (1 - pt(abs(t_classic), df_classic))
  
  # 95% confidence intervals
  lwr_welch <- est - se_welch * qt(p = 1 - (alpha / 2), df = df_welch)
  upr_welch <- est + se_welch * qt(p = 1 - (alpha / 2), df = df_welch)
  lwr_classic <- est - se_classic * qt(p = 1 - (alpha / 2), df = df_classic)
  upr_classic <- est + se_classic * qt(p = 1 - (alpha / 2), df = df_classic)
  
  # output
  data.frame(contrast = rep(rownames(contrast), times = 2),
             equal_var = rep(c("Assumed", "Not Assumed"), each = num_contrast),
             est = rep(est, times = 2),
             se = c(se_classic, se_welch),
             t = c(t_classic, t_welch),
             df = c(df_classic, df_welch),
             p = c(p_classic, p_welch),
             lwr = c(lwr_classic, lwr_welch),
             upr = c(upr_classic, upr_welch))
}

```

# simple demonstration first: paired t-test >> linear mixed effects model

# data

```{r}

# The group variable name may be misleading about the data: They represent measurements on 10 persons, not in groups.
kable(sleep)

```

## testing within-subjects effects using difefrent approaches

### using paired = TRUE

```{r}

t.test(extra ~ group, data = sleep, paired = TRUE)

```

### computing a difference score and testing difference score mean = 0

```{r}

# using dotproduct
t.test(with(sleep, cbind(extra[group == 1], extra[group == 2]) %*% c(-1, 1)))

# subtracting group 1 from group 2
t.test(with(sleep, extra[group == 1] * -1 + extra[group == 2] * 1))

```

### fitting linear mixed effects model

```{r}

# create new contrast code called drug, which represents the drug effect on sleep increase
sleep$drug <- with(sleep, ifelse(group == 1, -0.5, 0.5))

summary(lmer(extra ~ drug + (1 | ID), data = sleep, REML = TRUE))

```

# more complex situation with a between-subjects factor

## data

```{r}

# set random seed
set.seed(8888)

# id
id <- 1:50

# two treatment conditions, each size = 25
treatment <- rep(1:2, each = 25)

# labels in a separate column
treatment_lbl <- ifelse(treatment == 1, "control", "treatment")

# draws random numbers from true populations, depending on treatment value
response1 <- rnorm(n = rep(25, 2)[treatment], mean = c(4, 5)[treatment], sd = c(1, 1)[treatment])

# response 2 should be correlated r = 0.5 with response 1, but the mean should be slightly higher
response2 <- response1 * 0.5 + rnorm(n = 50, mean = c(1, 2)[treatment], sd = 1)

# save and print data
mydata <- tibble(id, treatment_lbl, treatment, response1, response2)
kable(mydata)

# restructure
lmydata <- gather(mydata, key = sequence_lbl, value = response, response1:response2) %>% 
  mutate(sequence = ifelse(sequence_lbl == "response1", -0.5, 0.5),
         treatment = ifelse(treatment == 1, -0.5, 0.5))

```

## treatment effect (between-subjects)

```{r}

# create new columns that averages the responses
mydata$response_avg <- with(mydata, (response1 + response2) / 2)

# using t-test
t.test(response_avg ~ treatment_lbl, data = mydata)

# using oneway function
oneway(dv = mydata$response_avg, group = mydata$treatment_lbl, contrast = list(treatment_lbl = c(-1, 1)), alpha = 0.05)

# using oneway function and dotproduct
oneway(dv = parse_number(as.matrix(mydata[, c("response1", "response2")]) %*% c(1, 1) / 2), group = mydata$treatment_lbl, contrast = list(treatment_lbl = c(-1, 1)), alpha = 0.05)

```

## sequence effect

```{r}

# create new columns that reflects difference between two response
mydata$response_diff <- with(mydata, -response1 + response2)

# using t-test on difference score, compare to mean = 0
with(mydata, t.test(response_diff))

# using paired t-test
t.test(response ~ sequence_lbl, data = lmydata, paired = TRUE)

# using oneway function
oneway(dv = mydata$response_diff, group = mydata$treatment_lbl, contrast = list(treatment_lbl = c(1, 1) / 2), alpha = 0.05)

# using oneway function and dotproduct
oneway(dv = parse_number(as.matrix(mydata[, c("response1", "response2")]) %*% c(-1, 1)), group = mydata$treatment_lbl, contrast = list(treatment_lbl = c(1, 1) / 2), alpha = 0.05)

```

## treatment x sequence effect

```{r}

# using independent samples welch t-test
t.test(response_diff ~ treatment_lbl, data = mydata)

# using oneway function
oneway(dv = mydata$response_diff, group = mydata$treatment_lbl, contrast = list(treatment_lbl = c(-1, 1)), alpha = 0.05)

# using oneway function and dotproduct
oneway(dv = parse_number(as.matrix(mydata[, c("response1", "response2")]) %*% c(-1, 1)), group = mydata$treatment_lbl, contrast = list(treatment_lbl = c(-1, 1)), alpha = 0.05)

```

# from Maxwell, Dalaney, and Kelley textbook

## data
> "For the hypothetical data contained in Table 14.8, a perceptual psychologist is interested in age differences ("young" and "old") in reaction time on a perceptual task. In addition, the psychologist is also interested in the effect of angle (zero degrees off center and eight degrees off center). The question of interest is to see if there are is a main effect of age, a main effect of angle, and an interaction between the two. Table 14.8 presents the same data that we analyzed in chapter 12 for 10 young participants and 10 old participants, except that for the moment we are only analyzing data from the 0 degree and 8 degree conditions of the angle factor."

```{r}

data("C14T8")

```

### add angle4

```{r}

# there is a slight difference between the textbook data and the data available in the package -- I fix that here
C14T8 <- C14T8 %>%
  mutate(subj = 1:20,
         Angle4 = c(510, 480, 630, 660, 660, 450, 600, 660, 660, 540, 570, 720, 540, 660, 570, 780, 690, 570, 750, 690),
         age = recode(Group, `1` = -0.5, `2` = 0.5),
         age_lbl = recode(Group, `1` = "young", `2` = "old")) %>% 
  select(subj, Group, age, age_lbl, Angle0, Angle4, Angle8)

# print
kable(C14T8)

```

### restructure

```{r}

lC14T8 <- C14T8 %>%
  gather(key = angle, value = rt, Angle0, Angle4, Angle8) %>%
  mutate(angle_num = str_sub(angle, start = nchar(angle), end = nchar(angle)) %>% parse_number(),
         angle_lbl = recode(angle_num, `0` = "0 degrees", `4` = "4 degrees", `8` = "8 degrees"),
         angle_linear = recode(angle_num, `0` = -0.5, `4` = 0, `8` = 0.5),
         angle_quadratic = recode(angle_num, `0` = 0.5, `4` = -1, `8` = 0.5))

```

## between-subjects contrast

### age contrast

```{r}

# c(1, 2, 3) %*% c(-1, 0, 1) means tke the dot product of the matrix c(1, 2, 3) and the matrix c(-1, 0, 1)
# this means multiply 1 by -1, 2 by 0, and 3 by 1, and add up those products
oneway(dv = parse_number(as.matrix(C14T8[, c("Angle0", "Angle4", "Angle8")]) %*% (c(1, 1, 1) / 3)), group = C14T8$Group, contrast = list(age = c(-1, 1)))

```

## within-subjects contrasts

#### linear angle contrast

```{r}

# c(1, 2, 3) %*% c(-1, 0, 1) means tke the dot product of the matrix c(1, 2, 3) and the matrix c(-1, 0, 1)
# this means multiply 1 by -1, 2 by 0, and 3 by 1, and add up those products
oneway(dv = parse_number(as.matrix(C14T8[, c("Angle0", "Angle4", "Angle8")]) %*% c(-1, 0, 1)), group = C14T8$Group, contrast = list(age = c(1, 1) / 2))

```

### quadratic angle contrast

```{r}

# c(1, 2, 3) %*% c(-1, 0, 1) means tke the dot product of the matrix c(1, 2, 3) and the matrix c(-1, 0, 1)
# this means multiply 1 by -1, 2 by 0, and 3 by 1, and add up those products
oneway(dv = parse_number(as.matrix(C14T8[, c("Angle0", "Angle4", "Angle8")]) %*% c(1, -2, 1) / 3), group = C14T8$Group, contrast = list(age = c(1, 1) / 2))

```

## mixed contrasts

### linear angle x age contrast

```{r}

# c(1, 2, 3) %*% c(-1, 0, 1) means tke the dot product of the matrix c(1, 2, 3) and the matrix c(-1, 0, 1)
# this means multiply 1 by -1, 2 by 0, and 3 by 1, and add up those products
oneway(dv = parse_number(as.matrix(C14T8[, c("Angle0", "Angle4", "Angle8")]) %*% c(-1, 0, 1)), group = C14T8$Group, contrast = list(age = c(-1, 1)))

```

### quadratic angle x age contrast

```{r}

# c(1, 2, 3) %*% c(-1, 0, 1) means tke the dot product of the matrix c(1, 2, 3) and the matrix c(-1, 0, 1)
# this means multiply 1 by -1, 2 by 0, and 3 by 1, and add up those products
oneway(dv = parse_number(as.matrix(C14T8[, c("Angle0", "Angle4", "Angle8")]) %*% c(1, -2, 1) / 3), group = C14T8$Group, contrast = list(age = c(-1, 1)))

```

## fit model with `lmer()`

```{r}

lmer01 <- lmer(rt ~ (angle_linear + angle_quadratic) * age + (1 + angle_linear + angle_quadratic | subj), data = lC14T8, REML = TRUE, control = lmerControl(check.nobs.vs.nRE = "ignore", check.nobs.vs.nlev = "ignore"))

```

### diagnostics

```{r}

plot(lmer01)
qqmath(ranef(lmer01, condVar = TRUE))
qqmath(lmer01)
dotplot(ranef(lmer01, condVar = TRUE))

```

### model results

```{r}

summary(lmer01)

```

# export data

```{r}

write_csv(lC14T8, path = "lC14T8.csv")

```
